---
title: "Medical appointment show up"
author: "Julius Rissanen"
date: "8.3.2017"
output: html_document
---
Abstract

'No Show appointments'-data provides information on people who have went to medical care and then missed or participated on the appointment. Not showing up is big problem as about 30% of people in dataset don't show up at the appointment. Dataset is really decently sized at 300 000 patients which provides good overview of the problem without small sample size problems. Logistic regression was chosen as statistical method to provide easily understandable model for intepreting the each variable's effect. Predictors which predicted increasing chance of missing the appointment were easily forecastable (alchoholism, smokers, saturday). Surprising predictor was before appointment sent SMS messages which didn't seem to have effect on people showing up. Logistic regression didn't provide good prediction accuracy in the end and thus more complex models would be needed if we want to get better at predicting who will miss their appointment.

```{r, message=FALSE, error=FALSE, warning=FALSE}
#needed libraries
library(ggplot2)
library(tidyverse)
library(reshape)
library(gridExtra)
library(stargazer)
library(magrittr)
library(boot)
```

#1. Medical appointment show up

##1.1 Description of the research question

My study looks to find what backgrounds in people can predict their probability of missing medical appointments.

##1.2 Data wrangling

Data wrangling file can be found from the github page:
https://github.com/JuliusRissanen/IODS-final/blob/master/data/create_noShow.R


##1.3 Description of the data

My data is open dataset from Kaggle. It has 300 000 patients and their status if they missed or didn't miss their medical appointment. Data can be downloaded from:
https://www.kaggle.com/joniarroba/noshowappointments

```{r}
#load dataset
noShow <- read.csv("data/noShow.csv", row.names = 1)

#fix RegistrationDate and AppoitmentDate to 'date' variables
noShow$RegistrationDate <- as.Date(noShow$RegistrationDate)
noShow$AppoitmentDate <- as.Date(noShow$AppoitmentDate)
#Fix binaries to logical
noShow$Diabetes <- as.logical(noShow$Diabetes)
noShow$Alchoholism <- as.logical(noShow$Alchoholism)
noShow$Hypertension <- as.logical(noShow$Hypertension)
noShow$Handicap <- as.logical(noShow$Handicap)
noShow$Smokes <- as.logical(noShow$Smokes)
noShow$Scholarship <- as.logical(noShow$Scholarship)
noShow$Tuberculosis <- as.logical(noShow$Tuberculosis)
noShow$Status <- as.logical(noShow$Status)

#Check structure of the data
str(noShow)
```

From the structure of the data we can se that it has 16 variables and 299 988 patients. 6 patients were deleted from the original dataset because of unreasonable values in age variable (<0) also only 6 people were on registered on sunday so they are also deleted from causing problems with DayOfTheWeek variable. Date and time based variables were cleaned to have only dates or hours. Handicap and SMS_reminder variables were recoded to binaries (from 0-4 scale Handicap and 0-2 amount of SMS send)

Variables are self explanatory from their names.

Continuous ones are Age and AwaitingTime. Awaiting time calculates how many days were between their reservation and medical appoitment.

Factor ones are Gender and DayOfTheWeek. Gender 2 level for male/female and DayOfTheWeek 7 level for weekdays

Binary ones are Diabetes, Alcholism, Hypertension, Smokes, Scholarship, Tuberculosis, handicap, SMS_reminder where 1 means person is e.g. diabetic ('Diabetes' variable).


##1.4 Data exploration

First we start with basic descriptive statistics
```{r}
summary(noShow)
```

Our average age is 37.81 (median 38.00), youngest is under 1 year old and oldest 113. Data is really skewed towards women them accounting around 60% of our whole dataset. All our binary (smokes, diabetes, smokes etc.) variables have under 10% of patients except for SMS-reminder which was sent to around 60% of patients. Average waiting is 13.84 days with median of 8 days. Registration hours for appointments are around noon (mean 11.27).

30% of people in the dataset missed the medical appointment.


###1.4.1 Age further exploration
```{r}
ggplot(noShow, aes(Age)) + geom_density(aes(color = as.factor(Status))) + theme_classic()
```

From the density plot we can see that Age is distributed thus that most of the mass is under 60 year olds. I would've guessed that older people were more active using medical services but in this dataset our intepretation needs to be done on the effect younger ones (unless we do separate analysis comparing older ones specifically). Also by plotting Age based on their Status-variable we can see that it looks like older ones are more likely to miss their appoitment.

###1.4.2 'Gender' further exploration

We will see graphically how gender affects arrival to medical appoitment.
```{r}
ggplot(noShow, aes(x=as.factor(Gender), fill= as.factor(Status))) + geom_bar(position ='fill') + xlab("Gender") + ylab("Percentage of no show up")
```

Gender doesn't seem to have clear difference in the affect of no show up.

###1.4.3 SMS-reminder further exploration

We will check basic graph on SMS-remind variable if it looks to have affect on showing up.

```{r}
ggplot(noShow, aes(x=as.factor(Sms_Reminder), fill = as.factor(Status))) + geom_bar(position = 'fill') + ylab("Percentage of no-show up") + xlab("SMS-reminder sent") + theme_classic()
```


Even closer difference.

###1.4.4 Other binary variables

We combine other (binary) variables to same plot as they seem to be similar in their distribution of no show ups (<10%) 

```{r}
keep_variables <- c("Diabetes", "Alchoholism", "Hypertension", "Handicap", "Smokes", "Scholarship", "Tuberculosis")
binaries <- select(noShow, one_of(keep_variables))
gather(binaries) %>% ggplot(aes(value), fill = Status) + facet_wrap("key", scales = "free") + geom_bar() + theme_bw()
```

And then next their distribution compared to Status

```{r}

Diabetes_graph <- ggplot(noShow, aes(x=Diabetes, fill=Status)) + geom_bar(position="fill")
  
Alchoholism_graph <- ggplot(noShow, aes(x=Alchoholism, fill=Status)) + geom_bar(position="fill")

Hypertension_graph <- ggplot(noShow, aes(x=Hypertension, fill=Status)) + geom_bar(position="fill")

Handicap_graph <- ggplot(noShow, aes(x=Handicap, fill=Status)) + geom_bar(position="fill")

Smokes_graph <- ggplot(noShow, aes(x=Smokes, fill=Status)) + geom_bar(position="fill")

Scholarship_graph <- ggplot(noShow, aes(x=Scholarship, fill=Status)) + geom_bar(position="fill")

Tuberculosis_graph <- ggplot(noShow, aes(x=Tuberculosis, fill=Status)) + geom_bar(position="fill")

grid.arrange(Diabetes_graph, Alchoholism_graph, Hypertension_graph, Handicap_graph, Smokes_graph, Scholarship_graph, Tuberculosis_graph, ncol=2, top='Binary variables')
```

Again, results seem very identical. Alchoholism and Tuberculosis seems to increase risk of not arriving. Diabetes and hypertension seems to increase the chance of coming to appointment. But we need further analysis to validate results and their statistical significance.


##1.5 Description of the statistical method

I'm going to use logistic regression to analyse dataset whether people arrive or do not arrive to medical appointments. Logistic regression is right tool for this type of problem because compared to linear regression we are working with binary dependent variable compared to continous (linear regression). Logistic regression is also preferrable against linear discriminant analysis because I'm more interested in probabilities how much each variable predicts the outcome compared to prediction and how variables "layout" in the linear discriminants in the biplot.

We will define linear model that can asses the probability of "success", in this case not showing up for medical appointment. We can define 'Odds' as p/(1-p), higher odds is corresponding to higher probability of success. To make it into logistic regression we can take log of odds ratio 
-> log(p/1-p) = a + bx + e which can be solved into 
P(Y=1) = 1/1+e^(-a-bx), which is our final form of logistic regression 

We can intepret exponents of estimated coefficients as odds ratios and thus give intepretations to estimates from statistical model.

##1.6 Analysis

###1.6.1 first model
We start by fitting our model to all variables except dates. We will also divide the dataset to training (80%) and test data for later validation.
```{r}
# number of rows 
n <- nrow(noShow)

# choose randomly 80% of the rows
rows <- sample(n,  size = n * 0.8)

# create train set
train <- noShow[rows,]

# create test set 
test <- noShow[-rows,]

# save the correct classes from test data
correct_classes <- test$Status

# remove the Status variable from test data
test <- select(test, -Status)

#We will run the logistic regression on the training data
m1 <- glm(Status ~ Age + Gender + DayOfTheWeek + Diabetes + Alchoholism + Hypertension + Handicap + Smokes + Scholarship + Tuberculosis + Sms_Reminder + AwaitingTime + RegistrationHour ,data = train, family = "binomial")
summary(m1)
```

All the variables except Gender, Diabetes and Handicap were statistically significant (+ sunday). 
Next we will get odds ratios and their confidence intervals with the help of R so we can do some intepretation

###1.6.2 ODDS RATIOS

```{r, message= FALSE, error=FALSE}
# compute odds ratios (OR)
OR <- coef(m1) %>% exp
# compute confidence intervals (CI)
CI <-confint(m1) %>% exp
# print out the odds ratios with their confidence intervals
cbind(OR, CI)
```

Strongest effects can be seen from the alcholism (1,40 OR), saturday (1,45 OR), smoking (1,27 OR) and scholarship (1,26 OR). Alchoholism, smoking and scholarship (medical appointment paid by someone else [public]) can be seen as indicators of the people with harder lifes. Saturday as an day-off might get people more likely to forget it and scholarship might.

On the other hand interesting finding is that sending SMS (0.90 OR) before appointment doesn't provide more benefits for medical centers as it's OR is one of the lowest. 

###1.6.3 second model and prediction

We will drop all the non statistically significant variables and run model again.

```{r}
m2 <- glm(Status ~ Age  + DayOfTheWeek + Alchoholism + Hypertension + Smokes + Scholarship  + Sms_Reminder + AwaitingTime + RegistrationHour ,data = train, family = "binomial")
summary(m2)
```

```{r}
#predict probabilities and add it into test data
probs <- predict(m2, test, type = "response")
test <- mutate(test, probability = probs)

# use the probabilities to make prediction of status
test <- mutate(test, prediction = probability > 0.5)


```
Confusion matrices for the predictions:
```{r}
#table the prediction versus correct Statuses
table(correct = as.logical(correct_classes), prediction = test$prediction)
table(correct = as.logical(correct_classes), prediction = test$prediction) %>% prop.table() %>% addmargins()
```

Next we calculate mean prediction error
```{r}
## training error
# define a loss function (mean prediction error)
loss_func <- function(class, prob) {
  n_wrong <- abs(class - prob) > 0.5
  mean(n_wrong)
}
loss_func(class = correct_classes, prob = test$probability)
```


And after that we do crossvalidated training error
```{r}
cv <- cv.glm(data = train, cost = loss_func, glmfit = m2, K = 10)
# average number of wrong predictions in the cross validation
cv$delta[1]
```

Our model doesn't provide really good prediction accuracy. Mean prediction error on test data is 0.303 and cross-validated is 0.304. From the earlier confusion matrix we can see that our model has a lot of problems identifying people who will not come to medical appointment. Only 102 were correctly classified to not coming compared 18050 false positives. False positives weren't as big of an problem. 

##1.7 Conclusion and discussion

By managing explain why someone misses the medical appointment would provide benefits for medical sector's  productivity with better planning.

Doing analysis with logistic regression allowed expressing the probabilitties (Odds ratios) about effect of each attainable variable. Factors which increase the probability of not coming weren't surprising but failure of SMS messages provided to be interesting. Further analysis on how to get people to not miss their appointments could provide useful information. 

Logistic regression didn't provide good prediction accuracy in the end. Some other more accurate but not as transparent methods could provide better prediction accuracy for the job compared to logistic regression which in the end is best used as relatively easily understandable statistical model.